{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-05-04 paper/main.tex
final fixes

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress,comma,round,numbers]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[all]{hypcap}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{lineno}
\linenumbers
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

To explore the effect of payoff memory, we only considered the simplest model of reciprocity. 
This leaves several possible directions for future work. 
For example, for both our analytical analysis and our simulations, we assumed that populations are {\it well-mixed}. 
As a result, any population member is equally likely to be a given player's last co-player. 
A similar analysis for structured populations would be desirable, as population structure can often promote cooperation without an explicit need for reciprocity~\citep{szabo:PR:2007,vanveelen:PNAS:2012}.
Similarly, we assumed that after identifying a suitable role model, players can easily imitate that role model's strategy. 
In the context of our study, this might be a particular simplification, because players with limited payoff memory may also have less accurate information to infer a co-player's strategy. 
 Finally, we have taken a player's payoff memory to be a given parameter of the model. 
 Instead, future models could explore how that payoff memory might co-evolve along with the players' strategies \citep[similar to previous work on the evolution of the players' feasible strategy sets, e.g.][]{Stewart2016,Schmid:PlosCB:2022,Murase:PLoSCompBio:2023a}. 
 Here, individuals might experience an interesting trade-off. 
 Better payoff-memory would help them to make better decisions, yet it might also come with higher cognitive costs. 
 For such a setup, our results suggest that in many instances, a modest amount of memory may be sufficient to achieve good outcomes.

@@

\noindent
{\bf Data accessibility.}
All data generated for this manuscript are openly available at:
\href{https://zenodo.org/records/10844104}{https://zenodo.org/records/10844104}.
The code for producing the results and generating the figures is available
on the online repository
\href{https://github.com/Nikoleta-v3/evolution-of-cooperation-with-limited-updating-payoff-memory}{evolution-of-cooperation-with-limited-updating-payoff-memory}.\\

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    The leftmost panels give a schematic overview of the two main scenarios we compare. 
    The two scenarios differ in how many past interactions individuals take into account when updating their strategy. 
    In the scenario with perfect payoff memory, individuals consider all their past interactions (against all population members, and taking every turn of each repeated game into account). 
    In the scenario with limited payoff memory, individuals only consider their very last interaction (against one specific population member, taking into account only one round of the repeated game). 
    The four panels  on the right side depict the outcome of evolutionary simulations for repeated games with either a low or a high benefit of cooperation. 
    Colors represent how often the respective region of the strategy space is visited over time. 
    In all four panels, two regions are visited particularly often. 
    One region corresponds to a neighborhood of \alld{} with $p\approx q\!\approx\!0$ (lower left corner). 
    The other region corresponds to a strip of conditionally cooperative strategies with $p\!\approx\! 1$ and $q$ satisfying the constraints \eqref{Eq:PerfectMemory} and \eqref{Eq:LimitedMemory}, respectively (lower right corner). 
    The resulting average cooperation rate depends on which of these two neighborhoods is visited more often.
    Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:expected_and_stochastic_for_donation}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=.75\textwidth]{cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf Evolution of direct reciprocity for different parameter values.} 
    To explore the robustness of our results, we have run simulations for different benefit values (left panels, {\it a}), and for different selection strengths (right panels, {\it b}). 
    In each case, we record the resulting average cooperation rate over the entire simulation (upper panels). 
    In addition we record the individuals' average generosity. 
    Here, we only take into account those residents with $p\! \approx\! 1$ and we compute
    the average of their cooperation probability~$q$. 
    These simulations suggest that perfect payoff memory consistently leads to more cooperation and more generosity. 
    Unless explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
    Simulations are run for $T\!=\!5\times 10^7$ time steps, and each point represents a single simulation run.}
    \label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies. 
  From left to right, we consider the following cases.
  ({\it i}) Updating occurs based on expected payoffs (perfect memory), 
  ({\it ii}) it occurs based on the last round of one interaction (limited memory), 
  ({\it iii}) based on the last round of two interactions,
  ({\it iv}) based on the last two rounds of one interaction,
  ({\it v}) based on the last two rounds of two interactions, and
  ({\it vi}) based on the average payoff of one interaction.
  Again, simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!3$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate. 
  However, when individuals take into account at least two past interactions -- cases ({\it iii}) to ({\it vi}) -- evolving cooperation rates are close to this optimum. 
  Simulations are run for $T\!=\!10^7$ time steps, based on the parameters $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress,comma,round,numbers]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[all]{hypcap}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{lineno}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

To explore the effect of payoff memory, we only considered the simplest model of reciprocity. 
This leaves several possible directions for future work. 
For example, for both our analytical analysis and our simulations, we assumed that populations are {\it well-mixed}. 
As a result, any population member is equally likely to be a given player's last co-player. 
A similar analysis for structured populations would be desirable, as population structure can often promote cooperation without an explicit need for reciprocity~\citep{szabo:PR:2007,vanveelen:PNAS:2012}.
Similarly, we assumed that after identifying a suitable role model, players can
easily imitate that role model's strategy. Especially in repeated games, this
assumption is less innocuous than it may appear. After all, a player can
typically only observe the co-players' actions, but not their underlying
strategies. This player may only attempt to infer the co-players' strategies
from their past reactions. Once players only have a restricted recollection of
the past, this task seems particularly difficult. In the context of our study,
this might be a particular simplification, because players with limited payoff
memory may also have less accurate information to infer a co-player's strategy.
Finally, we have taken a player's payoff memory to be a given parameter of the model. 
Instead, future models could explore how that payoff memory might co-evolve along with the players' strategies \citep[similar to previous work on the evolution of the players' feasible strategy sets, e.g.][]{Stewart2016,Schmid:PlosCB:2022,Murase:PLoSCompBio:2023a}. 
Here, individuals might experience an interesting trade-off. 
Better payoff-memory would help them to make better decisions, yet it might also come with higher cognitive costs. 
For such a setup, our results suggest that in many instances, a modest amount of memory may be sufficient to achieve good outcomes.

@@

\noindent
{\bf Data accessibility.}
All data generated for this manuscript are openly available at:
\href{https://zenodo.org/records/10844104}{https://zenodo.org/records/10844104}.
The code for producing the results and generating the figures is available
on the online GitHub repository
\href{https://github.com/Nikoleta-v3/evolution-of-cooperation-with-limited-updating-payoff-memory}{evolution-of-cooperation-with-limited-updating-payoff-memory}.\\

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    The leftmost panels give a schematic overview of the two main scenarios we compare. 
    The two scenarios differ in how many past interactions individuals take into account when updating their strategy. 
    In the scenario with perfect payoff memory, individuals consider all their past interactions (against all population members, and taking every turn of each repeated game into account). 
    In the scenario with limited payoff memory, individuals only consider their very last interaction (against one specific population member, taking into account only one round of the repeated game). 
    The four panels  on the right side depict the outcome of evolutionary simulations for repeated games with either a low or a high benefit of cooperation. 
    Colors represent how often the respective region of the strategy space is visited over time. 
    In all four panels, two regions are visited particularly often. 
    One region corresponds to a neighborhood of \alld{} with $p\approx q\!\approx\!0$ (lower left corner). 
    The other region corresponds to a strip of conditionally cooperative strategies with $p\!\approx\! 1$ and $q$ satisfying the constraints \eqref{Eq:PerfectMemory} and \eqref{Eq:LimitedMemory}, respectively (lower right corner). 
    The resulting average cooperation rate depends on which of these two neighborhoods is visited more often.
    Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:expected_and_stochastic_for_donation}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=.75\textwidth]{static/cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf Evolution of direct reciprocity for different parameter values.} 
    To explore the robustness of our results, we have run simulations for different benefit values (left panels, {\it a}), and for different selection strengths (right panels, {\it b}). 
    In each case, we record the resulting average cooperation rate over the entire simulation (upper panels). 
    In addition we record the individuals' average generosity. 
    Here, we only take into account those residents with $p\! \approx\! 1$ and we compute
    the average of their cooperation probability~$q$. 
    These simulations suggest that perfect payoff memory consistently leads to more cooperation and more generosity. 
    Unless explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
    Simulations are run for $T\!=\!5\times 10^7$ time steps, and each point represents a single simulation run.}
    \label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies. 
  From left to right, we consider the following cases.
  ({\it i}) Updating occurs based on expected payoffs (perfect memory), 
  ({\it ii}) it occurs based on the last round of one interaction (limited memory), 
  ({\it iii}) based on the last round of two interactions,
  ({\it iv}) based on the last two rounds of one interaction,
  ({\it v}) based on the last two rounds of two interactions, and
  ({\it vi}) based on the average payoff of one interaction.
  Again, simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!3$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate. 
  However, when individuals take into account at least two past interactions -- cases ({\it iii}) to ({\it vi}) -- evolving cooperation rates are close to this optimum. 
  Simulations are run for $T\!=\!10^7$ time steps, based on the parameters $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

### 2024-05-04 paper/si.tex
final fixes

\title{{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory}}
\date{}
\author{Nikoleta E. Glynatsi, Alex McAvoy, Christian Hilbe}

>>>>

\title{{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory}}
\date{}
\author{\parbox[c]{16cm}{\centering \onehalfspacing 
Nikoleta E. Glynatsi$^1$,  Alex McAvoy$^{2,3, \dagger}$, Christian Hilbe$^{1, \dagger}$\\ \quad \\
$^{\rm 1}$Max Planck Research Group on the Dynamics of Social Behavior,\\ Max Planck Institute for Evolutionary Biology, Pl\"{o}n, Germany \\
$^{\rm 2}$School of Data Science and Society, University of North Carolina at Chapel Hill,\\ Chapel Hill, NC 27599 \\
$^{\rm 3}$Department of Mathematics, University of North Carolina at Chapel Hill,\\ Chapel Hill, NC 27599 \\
$^{\rm \dagger}$ A.M. and C.H. contributed equally to this work.}}
